---
title: "Data Analysis #2 Version 2 (75 points total)"
author: "Rico, Christopher"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
# library(rockchalk) # base R code replaces requirement for this package
library(car)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

##### Section 1: (5 points)

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}

#get volume and ratio back into dataset
mydata$VOLUME = mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO = mydata$SHUCK / mydata$VOLUME

#generate histogram and qqplot of ratio
hist(mydata$RATIO, main = "Histogram of Shuck/Vol Ratio")
qqnorm(mydata$RATIO, main = "Q-Q Plot of Shuck/Vol Ratio")
qqline(mydata$RATIO)

#calculate skewness and kurtosis using base r functions (as noted above, rockchalk would not be loaded)
skewness(mydata$RATIO)
kurtosis(mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}

mydata$L_RATIO = log10(mydata$RATIO)

#boxplots and QQ plot of logarithmic ratio
hist(mydata$L_RATIO, main = "Histogram of Shuck/Vol Ratio")
qqnorm(mydata$L_RATIO, main = "Q-Q Plot of Shuck/Vol Ratio")
qqline(mydata$L_RATIO)

skewness(mydata$L_RATIO)
kurtosis(mydata$L_RATIO)


boxplot(mydata$L_RATIO ~ mydata$CLASS, 
        main = "Boxplots of Log Ratio x Class",
        xlab = "Log Ratio Dist x Class",
        ylab = "Freq")

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}

bartlett.test(RATIO ~ CLASS, data = mydata)
bartlett.test(L_RATIO ~ CLASS, data = mydata)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: (Enter your answer here.)***
L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes. This is shown with less skew in the histogram, more evenly distributed outliers in the QQ plot and boxplots, less skewness in the QQ plot, and failing to reject the null hypothesis with the Bartlett test of homogeneity of variances.
This is due to the 'muting' nature of taking the log value of a number. Logarithms of series of numbers have less graphical variation than the raw number, because logarithms are a display of orders of magnitude.

##### Section 2 (10 points) ###############################

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

summary(aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: (Enter your answer here.)***
The interaction term had little to no effect and was not significant. However, the main effects of CLASS and SEX are statistically significant, the interaction between the two is not.

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}

TukeyHSD( aov(L_RATIO ~ CLASS + SEX, data = mydata) )

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: (Enter your answer here.)***
Yes! We have rejected the null hypothesis that infant ratio vs male/female ratio is the same. However, we did not reject the null hypothesis that male ratio are significantly different than female ratio. This suggests that males and females can be combined to a single 'adult' category.

######  Section 3: (10 points) ##################

(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}

par(mfrow = c(1, 2))

hist(mydata$VOLUME[mydata$TYPE == 'I'],
     xlab = "Volume",
     main = "Infant Volumes")
hist(mydata$VOLUME[mydata$TYPE == 'ADULT'],
     xlab = "Volume",
     main = "Adult Volumes")
```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: (Enter your answer here.)***
The infant distribution is much more heavily right skewed compared to the more normal adult distribution. There appears to be a divide at a volume of around 300 -- most adults are larger than 300, and most infants smaller than. This could point to volume being a good metric to split based on.

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}

#add columns for base 10 logarithms of shuck and volume
mydata$L_SHUCK = log10(mydata$SHUCK)
mydata$L_VOLUME = log10(mydata$VOLUME)

#plot scatterplots for shuck vs volume and fill by CLASS and TYPE 
grid.arrange(
  ggplot(mydata, aes(x=VOLUME, y=SHUCK, color = CLASS)) + geom_point(aes(fill = CLASS)),
  ggplot(mydata, aes(x=L_VOLUME, y=L_SHUCK, color =CLASS)) + geom_point(aes(fill = CLASS)),
  
  ggplot(mydata, aes(x=VOLUME, y=SHUCK, color = TYPE)) + geom_point(aes(fill = TYPE)),
  ggplot(mydata, aes(x=L_VOLUME, y=L_SHUCK, color = TYPE)) + geom_point(aes(fill = TYPE))
)

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: (Enter your answer here.)***
VOLUME and SHUCK plots overlap pretty heavily, making it difficult to distinguish clear lines between CLASS or TYPE. Log-transforming these measures helps clarify divisions between A1 and infant abalones compared to the other classes and type. The infant Type has a large group at (<1.75, <0.75) and adult type is almost completely above and to the right of these points. Log-transforming these measurements seems as if it helps tighten up the relationships to be more linear and distinct, which would arguably allow for much better fit to a linear regression model. That is super useful and neat!

######   Section 4: (5 points) ###################################

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))
```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model = lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: (Enter your answer here.)***
The Estimate coefficients suggest a stronger negative correlation in L_SHUCK for the higher the class. This combined with the previous scatterplots suggests that L_SHUCK increases more readily at the lower classes and then either flattens out or even decreases on average. Graphically, this trend is not as apparent for the L_SHUCK by class scatterplots -- although logically it makes sense, as L_SHUCK is graphed by orders of magnitude of SHUCK. We can see this trend more easily if we look at the scatterplot for SHUCK: notice how class A5 seems to trend with a more negative slope than do than the other adult classes. 

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: (Enter your answer here.)***
Type is less important than most of the upper classes, and much less important than L_VOLUME (just look at the estimate coefficients in the summary). I would not choose to use type to predict L_SHUCK.
-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

###### Section 5: (5 points) #################################

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}

#compute both skewness and kurtosis
skewness(model$residuals)
kurtosis(model$residuals)

#generate histogram and QQplot
hist(model$residuals, 
     main = "Residuals",
     xlab = "Residuals")

qqnorm(model$residuals, 
       main = "Q-Q Plot")
qqline(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}

#add in residuals to mydata
mydata$RESIDUALS = model$residuals

#plot scatterplots for shuck vs volume and fill by CLASS and TYPE 
grid.arrange(
  ggplot(mydata, aes(x=L_VOLUME, y=RESIDUALS, color = CLASS)) + geom_point(aes(fill = CLASS)),
  ggplot(mydata, aes(x=L_VOLUME, y=RESIDUALS, color =TYPE)) + geom_point(aes(fill = TYPE)),
  
  ggplot(mydata, aes(x=CLASS, y=RESIDUALS, color = CLASS)) + geom_boxplot(aes(fill = CLASS)),
  ggplot(mydata, aes(x=TYPE, y=RESIDUALS, color = TYPE)) + geom_boxplot(aes(fill = TYPE))
)

#perform a bartlett test
bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: (Enter your answer here.)***
The histogram and Q-Q plot in 5a show us that the residuals for the model are roughly normal, with most residual values falling around 0. This is a good indicator that the majority of the variability in the data is explained by the model (as opposed to a skewed histogram, which would indicate that much of the variability in the data is not explained by the model). We can see similar trends in the charts in 5b, which show the data somewhat 'blobbed' around residual values of 0.
This tells us that L_VOLUME and VOLUME both are potentially useful for making harvesting decisions, as volume is a decently accurate predictor of both class and type.
-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

#### Section 6: (5 points) ########################

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}

#generate a plot 
ggplot() + 
  #with a line for the volume.value vs infants and adults
  geom_line( aes(volume.value, prop.infants), color = "orange" ) +
  geom_line( aes(volume.value, prop.adults), color = "blue" ) +
  
  #and a line for 50% split volume.value for each
  geom_vline(xintercept = split.infants) +
  #with a label on the line
  annotate( "text", label = paste(round(split.infants, 2), "\nInfants"),
           x = split.infants + 56, y = 0.5, color = "orange" ) +
  geom_vline(xintercept = split.adults) +
  annotate( "text", label = paste(round(split.adults, 2), "\nAdults"),
           x = split.adults + 75, y = 0.5, color = "blue" ) +
  labs(x = "Volume", y = "Proportion",
       title = "Proportion of Adults and Infants protected")

```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: (Enter your answer here.)***
There is a nice, large distinction between the two values (one is more than double the other), which would help in reducing the potential for accidentally harvesting infants or not harvesting adults. This is a good thing, because the clearer the distinction between the two groups, the easier it will be for harvesters in the field to tell the difference between the two populations.

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###### Section 7: (10 points)  #######################

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

diff = (1 - prop.adults) - (1 - prop.infants)

ggplot() +
  geom_line( aes(volume.value, diff) ) +
  
  geom_vline(xintercept = split.infants) +
  annotate("text", label = paste('infant volume =', round(split.infants, 2)), 
           x = split.infants + 20, y = 0.2, angle = 90) +
  geom_vline(xintercept = split.adults) +
  annotate("text", label = paste('adult volume =', round(split.adults, 2)), 
           x = split.adults + 20, y = 0.2, angle = 90) +
  
  labs(x = "Volume", y = "Diff in Proportions Harvested", 
     title = "Diff in Harvest Proportions")
```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}

cutoff1 = volume.value[ which.max(smooth.difference) ]

ggplot() +
  geom_line( aes(volume.value, diff), color = "orange" ) + 
  geom_line( aes(volume.value, smooth.difference), color = "blue" ) +
  
  geom_vline( xintercept = volume.value[cutoff1] ) +
  annotate( "text", label = paste('peak vol = ', round(cutoff1, 3) ),
           x = cutoff1 + 10, y = 0.35) +
  
  labs(x = "Volume", y = "Diff in Proportions Harvested",
       title = "Diff in Proportions Harvested")

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
(1 - prop.infants)[which.max(smooth.difference)]

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

######  Section 8: (10 points)  ###################

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}

cutoff2 = volume.value[ volume.value > max(mydata[mydata$CLASS == "A1" &
                                           mydata$TYPE == "I", "VOLUME"]) ][1] # [1] 206.786

(1 - prop.adults)[which(volume.value == cutoff2)]
(1 - prop.infants)[which(volume.value == cutoff2)]

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}

cutoff3 = volume.value[ which.min(abs(prop.adults - (1-prop.infants))) ] # [1] 237.6391

(1 - prop.adults)[which(volume.value == cutoff3)]
(1 - prop.infants)[which(volume.value == cutoff3)]

```


##### Section 9: (5 points) ###########

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}

cutoffs = which(volume.value %in% c(cutoff1, cutoff2, cutoff3))

ggplot( mapping = aes((1 - prop.infants), (1 - prop.adults)) ) +
  geom_line(color = "blue") +
  geom_point(mapping = aes( (1 - prop.infants[cutoffs]), (1 - prop.adults[cutoffs])) ) +
  
  #stick some labels on there
  annotate("text", label = paste('Max. difference \nvol =', round(cutoff1, 1)), 
           x = 0.1, y = 0.82) +
  annotate("text", label = paste('0 A1 infants \nvol =', round(cutoff2, 1)), 
           x = 0.29, y = 0.69) +
  annotate("text", label = paste('Equal harvest/conserve \nvol =', round(cutoff3, 1)), 
           x = 0.3, y = 0.95) +
  
  labs(x = "Infant Harvest Proportion", y = "Adult harvest proportion",
      title = "ROC curve of adult/infant harvest proportions")

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}

auc(x = (1 - prop.infants), y = (1 - prop.adults))

```


##### Section 10: (10 points) ###################

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	

strat = c("0 A1 Infants", "Equal Error", "Max Diff")
cuts = round( c(cutoff2, cutoff3, cutoff1), 1)

#calculate true and false positive rates
truePos = round( (1 - prop.adults)[cutoffs], 3)
falsePos = round( (1 - prop.infants)[cutoffs], 3)
#as well as yield
yield = round( (truePos * total.adults + falsePos * total.infants)/
        (total.adults + total.infants), 3)

#generate matrix of values
ROCtable = matrix( c(strat, cuts, truePos, falsePos, yield), ncol = 5, byrow = FALSE )
#and column names
colnames(ROCtable) = c("Strategy", "Volume", "TPR", "FPR", "Yield")

ROCtable = as.table(ROCtable)
ROCtable

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: (Enter your answer here.)***
There is a wide range of possible volumes represented in the cutoffs. However, the true positive rate, false positive rate, and harvest proportion are actually all three fairly close to one another (all within about 10%). 
- Harvesting 0 A1 infants leads to the greatest true positive rate (and also the greatest false positive rate), as well as the greatest harvest proportion of all. This could be called the least conservative strategy.
- Harvesting with an 'Equal Error' strategy is the most moderate approach of the three with regards to TPR, FPR and HP. 
- Harvesting with the 'Maximum Difference' strategy is by far the most conservative approach, and gives the smallest TPR, FPR, and HP.


**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: (Enter your answer here.)***
Although I have a biology degree, I am certainly not a marine biologist. I would not feel qualified to make a determination about which cutoff is the best choice for balancing abalone conservation with abalone hunting. The best that I could do given the opportunity would be to present these analyses and explain the meaning of each cutoff value. It would require someone with greater knowedge of abalone population dynamics to give real-world context to the numbers, and decide how best to implement some sort of harvest cutoff policy.
I would of course qualify the model we generated to be relevant solely to the abalone population from which the sample data were taken. Different species of abalone growing in a different environment may have a completely different morphology, which would render our model useless for predicting abalone maturity based on volume.
However, if I was the sole decider of a harvest policy, I would aim towards the more conservative end of the spectrum to protect infant abalone -- maybe splitting the difference between the Maximum Difference and Equal Error strategies would be a good balance. Any abalone with a volume greater than 250 would be fair game to harvest.
Unfortunately, not much information was given to us about the dataset when we started -- I would request that the data be as conclusive as possible to as many different abaolone populations as possible to give a more general model to use.
