---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Rico, Christopher"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections ##########################

##### Section 1: (6 points) Summarizing the data.

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (Enter your answer here.)***
Nominal variables: Sex
Ordinal variables: Class, Rings
Ratio variables: Length, Diam, Height, Whole, Shuck, Volume, Ratio (I think these are ratio variables because they have a clearly defined zero value and no negatives)

Honestly, I have a really hard time determining potential outliers from summary tables. I can see, however, that most of the data are roughly normally distributed, which is not a surprise.


(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
#make the table object
sexClassTable = table(mydata$SEX, mydata$CLASS)

#add in the sum margins
addmargins( sexClassTable )

barplot(sexClassTable, 
        main ="Abalone Sex Distribution by Class", 
        xlab = "Class",
        legend = rownames(sexClassTable))
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (Enter your answer here.)***
Male and female abalones are fairly evenly distributed regardless of the class, with perhaps a slightly greater proportion of females relative to males in classses A1, A2, and A3. 
However, abalones in classes A1 and A2 are mostly infants, which we can take to mean that they are unsexed as of yet. Infant to Male/Female ratio drops quickly as class number increases, until in classes A4 and A5, infants are a much smaller proportion of the sexual makeup relative to male or female abalones. Of course, this makes sense: class is a classification of age, and the greater the class, the older the abalone. It appears that older abalones are much more likely to have a determined sex than younger abalones.
Also, we can see that alabalones, regardless of sex, are less populous in both the very young (A1) and very old (A4, A5) age classes. The most populous class is A3 -- young adult abalone, if you will.


(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
#set the 'random' seed generator
set.seed(123)

#get the sample out of mydata
work = mydata[ sample(1:nrow(mydata), 200, replace = FALSE), ]

#make the plots of work
plot( work[, 2:6] )
```

-----

##### Section 2: (5 points) Summarizing the data using graphics.

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$VOLUME, mydata$WHOLE,
     main = "Abalone Whole weight (g) by Volume (cm^3)",
     xlab = "Volume (cm^3)", ylab = "Whole weight (g)",
     col = 1:5,
     pch = 19
     )

legend(x = 100, y = 300, 
       c("A1", "A2", "A3", "A4", "A5"), 
       col = 1:5,
       pch = 19
       )
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK,
     main = "Abalone Shuck weight (g) by Whole weight (g)",
     xlab = "Shuck (g)", ylab = "Whole weight (g)",
     col = 6:10,
     pch = 19
     )

legend(x = 10, y = 100, 
       c("A1", "A2", "A3", "A4", "A5"), 
       col = 6:10,
       pch = 19
       )

#graph the line, with the slope as the max ratio of SHUCK/WHOLE
abline(a = 0, b = max(mydata$SHUCK/mydata$WHOLE))
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (Enter your answer here.)***
The variability in this plot differs from that in plot (a) somewhat. We can see that there's a somewhat tighter fit to the points in this plot, since while the two charts look about the same, the scale on this chart is smaller than that in plot (a). This leads me to believe that the weight of the shell is not the source of as much variability in the weight data as the shuck weight is. To be honest, I have a really hard time teasing apart the different age classes from these charts, so I am not confident enough to talk about the potential contributions of the age class distributions.
-----

##### Section 3: (8 points) Getting insights about the data using graphs.

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
#subset mydata into different sexes
maleSub = subset(mydata, mydata$SEX == "M")
femSub = subset(mydata, mydata$SEX == "F")
infSub = subset(mydata, mydata$SEX == "I")

par( mfrow = c(3,3) )

#plot histograms
hist(maleSub$RATIO, col = "red")
hist(femSub$RATIO, col = "green")
hist(infSub$RATIO, col = "blue")

#plot boxplots
boxplot(maleSub$RATIO, col = "red")
boxplot(femSub$RATIO, col = "green")
boxplot(infSub$RATIO, col = "blue")

#plot Q-Q plots
qqnorm(maleSub$RATIO, col = "red")
qqline(maleSub$RATIO)
qqnorm(femSub$RATIO, col = "green")
qqline(femSub$RATIO)
qqnorm(infSub$RATIO, col = "blue")
qqline(infSub$RATIO)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Enter your answer here.)***
All three sub-populations (Male, Female, and Infant) have Ratio data that are right skewed, and, for the most part, look fairly simlilar to one another. However, the infant population has a somewhat greater number of high-value outliers than do the male or female populations. I would like to see these charts broken down by class to see if the large infant outliers are the small numbers of infants who are in the A4 and A5 classes, as I would expect these to haave a larger shuck weight and therefore a large ratio.

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
#identify boxplot outliers for each sex, filter each dataset to only lines with outliers
maleOut = subset(maleSub, maleSub$RATIO %in% boxplot.stats(maleSub$RATIO)$out)
femOut = subset(femSub, femSub$RATIO %in% boxplot.stats(femSub$RATIO)$out )
infOut = subset(infSub, infSub$RATIO %in% boxplot.stats(infSub$RATIO)$out )

#and print the results
kable(maleOut)
kable(femOut)
kable(infOut)
#
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (Enter your answer here.)***
Well, for one thing, I was wrong about my expectation above in (3)(a): the infant outliers come almost exclusively from the A1 and A2 class, which is the opposite of what I expected! The male and female outliers, on the other hand, come all from classes A3, A4, and A5, and although most of them are outliers on the large side of the normal distribution (which I would have expected after looking at the boxplot earlier), female population only has a single small outlier. Weird!
-----

##### Section 4: (8 points) Getting insights about possible predictors.

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par( mfrow = c(2,2) )

boxplot(mydata$VOLUME~mydata$CLASS, 
        xlab = "Class", ylab = "Volume (cm^3)")
boxplot(mydata$WHOLE~mydata$CLASS, 
        xlab = "Class", ylab = "WHOLE (g)")

plot(mydata$RINGS, mydata$VOLUME,
     xlab = "RINGS", ylab = "VOLUME")
plot(mydata$RINGS, mydata$WHOLE,
     xlab = "RINGS", ylab = "WHOLE")
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (Enter your answer here.)***
Honestly, I do not think either of these variables would be particularly good predictors of age. Volume and Whole weight appear to correlate a little bit more tightlty with class than they do with rings. However, looking at the scatterplots, we can see that there is a very, very noisy relationship that probably has a pretty abysmal regression score. This means that it would be really tough to use these variables to get accurate predictions, becuase there are so many potential outputs for a given input. This is not what we want to see! Perhaps there are a few other variables that would allow us to tighten up the ability to predict and show a stronger correlation. Still though, we are not positive that any of these relationships are causal! I suppose that the point of a predictor is not necessarily to determine a causal relationship but to discover how well a value can be predicted from other values.
-----

##### Section 5: (12 points) Getting insights regarding different groups in the data.

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
#make aggregate tables
agVol = aggregate(VOLUME~SEX+CLASS, data = mydata, mean)
agShuck = aggregate(SHUCK~SEX+CLASS, data = mydata, mean)
agRatio = aggregate(RATIO~SEX+CLASS, data = mydata, mean)

#matrix-ify and name the volume aggregate
volume_aggregate = matrix(agVol$VOLUME, nrow = 3, ncol = 5)
rownames(volume_aggregate) <- c("Female", "Infant", "Male")
colnames(volume_aggregate) <- c("A1","A2","A3","A4","A5")

#matrix-ify and name the shuck aggregate
shuck_aggregate = matrix(agShuck$SHUCK, nrow = 3, ncol = 5)
rownames(shuck_aggregate) <- c("Female", "Infant", "Male")
colnames(shuck_aggregate) <- c("A1","A2","A3","A4","A5")

#matrix-ify and name the ratio aggregate
ratio_aggregate = matrix(agRatio$RATIO, nrow = 3, ncol = 5)
rownames(ratio_aggregate) <- c("Female", "Infant", "Male")
colnames(ratio_aggregate) <- c("A1","A2","A3","A4","A5")

#present three matrices
volume_aggregate
shuck_aggregate
ratio_aggregate
```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

ggplot(data = agRatio, aes(x = CLASS, y = RATIO, group = SEX, color = SEX)) +
      geom_line() + geom_point()+
      ggtitle("Plot showing Ratio by Class")

ggplot(data = agVol, aes(x = CLASS, y = VOLUME, group = SEX, color = SEX)) +
      geom_line() + geom_point()+
      ggtitle("Plot showing volume by Class")

ggplot(data = agShuck, aes(x = CLASS, y = SHUCK, group = SEX, color = SEX)) +
      geom_line() + geom_point()+
      ggtitle("Plot showing sShuck weight by Class")
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***
Why do abaolone shell weights increase relative to the shuck weights with age? That seems like such a strange process: I would expect abalones to grow meatier with age! Maybe their shells grow thicker with age to protect them more thoroughly.
Abalone volume growing with age definitely makes sense, because most organisms grow as they get older! Same goes for the shuck weight. I fully expect infants -- regardless of class -- to have a smaller shuck weight than either male or female abalone. Maybe there's a relationship between weight and sex determination?

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
adultSub = subset(mydata, mydata$SEX == "M" | mydata$SEX == "F")
infantSub = subset(mydata, mydata$SEX == "I")

par( mfrow = c(2,2) )

boxplot(adultSub$VOLUME~adultSub$RINGS, 
        xlab = "Rings", ylab = "Volume (cm^3)",
        main = "Adult Volume | Rings")


boxplot(infantSub$VOLUME~infantSub$RINGS, 
        xlab = "Rings", ylab = "Volume (cm^3)",
        main = "Infant Volume | Rings")


boxplot(adultSub$WHOLE~adultSub$RINGS, 
        xlab = "Rings", ylab = "Whole Weight (g)",
        main = "Adult Whole Weight | Rings")


boxplot(infantSub$WHOLE~infantSub$RINGS, 
        xlab = "Rings", ylab = "Whole Weight (g)",
        main = "Infant Whole Weight | Rings")


```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (Enter your answer here.)***
Infants are, as expected, mostly smaller in weight and possess fewer rings than do adults. However, infant abalone with a large number of rings seem to have a pretty large variability in their whole weight! Adults with a moderate number of rings seem to have an enormous variability in whole weight.

-----

##### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).

**Conclusions**

**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Enter your answer here.)***
Let's put statistics aside for a moment and just talk about basic biology: a big reason that this study failed was that the researchers failed to take into account the nature of phenotypic expression. Every organism will express plasticity in its phenotype depending on the conditions it lives in. If, for example, an organism has access to greater nutrients than other organisms of the same species, it will of course grow to be larger and heavier. Humans, when exposed to UV rays, tend to develop more melanin in their skin. 
The researchers did not take these external variables into account, and so of course the data they collected were not predictive of age: the sample of the population was too noisy! To get better predictability, the researchers might consider breaking out the samples into different geographical areas and trying to ascertain size as a predictor of age at a more granular level. This would help to remove some of the variability in abalone habitat, which I would imagine to be a big contributor to variability in size and weight. 


**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (Enter your answer here.)***
I would have a lot of questions regarding the sampling methodology: what proportion of the population is this sample? How were samples chosen? Do we know population characteristics for the population so we can compare the sample's characteristics to those of the population? Does the sample size relative to the sample variability give us statistical significance for any sort of scores we may need to generate?


**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Enter your answer here.)***
Observational studies certainly have value, but the question brings up an excellent point: we can usually not determine a causal relationship by performing observational studies alone. While conclusions about one trait's values ability to predict others can certainly be determined and characterized by an observational study, we cannot truly uncover a causal relationship when we solely observe. The nature of an observational study is simply to observe and collect data passively, without changing any inputs. Because of this passivity, we are not able to see that change in one value *causes* change in another value. This does limit the size and scope of the conclustions we are able to make.